---
title: "Data Exploration"
author: "Your Name"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)

# Load required libraries
library(dplyr)
library(ggplot2)
library(plotly)
library(DT)
library(corrplot)
library(here)

# Source project functions
source(here("R", "utils.R"))
source(here("R", "data_processing.R"))

# Load configuration
config <- load_config()
setup_logging(config)
```

## Project Overview

Brief description of the project goals and the dataset being analyzed.

## Data Loading

```{r load-data}
# Load raw data
data_path <- here(config$data$raw_path, "dataset.csv")
raw_data <- load_raw_data(data_path, config)

# Display basic information
cat("Dataset dimensions:", nrow(raw_data), "rows ×", ncol(raw_data), "columns\n")
```

## Data Structure

```{r data-structure}
# Display data structure
str(raw_data)

# Show first few rows
DT::datatable(
  head(raw_data, 100),
  options = list(scrollX = TRUE, pageLength = 10),
  caption = "First 100 rows of the dataset"
)
```

## Summary Statistics

```{r summary-stats}
# Numerical summaries
numerical_cols %
    summarise_all(~ paste("Unique values:", n_distinct(.))) %>%
    tidyr::pivot_longer(everything(), names_to = "Variable", values_to = "Info") %>%
    knitr::kable()
}
```

## Missing Data Analysis

```{r missing-data}
# Calculate missing data percentages
missing_data %
  summarise_all(~ sum(is.na(.)) / length(.) * 100) %>%
  tidyr::pivot_longer(everything(), names_to = "Variable", values_to = "Missing_Percent") %>%
  arrange(desc(Missing_Percent))

# Visualize missing data
missing_plot %
  filter(Missing_Percent > 0) %>%
  ggplot(aes(x = reorder(Variable, Missing_Percent), y = Missing_Percent)) +
  geom_col(fill = "steelblue", alpha = 0.7) +
  coord_flip() +
  labs(
    title = "Missing Data by Variable",
    x = "Variables",
    y = "Missing Percentage (%)"
  ) +
  theme_minimal()

ggplotly(missing_plot)

# Display missing data table
DT::datatable(
  missing_data,
  options = list(pageLength = 15),
  caption = "Missing data summary"
) %>%
  DT::formatRound("Missing_Percent", digits = 2)
```

## Target Variable Analysis

```{r target-analysis}
target_col %
    count(!!sym(target_col)) %>%
    mutate(percentage = n / sum(n) * 100)
  
  print(target_summary)
  
  # Visualize target distribution
  target_plot %
    ggplot(aes(x = !!sym(target_col))) +
    geom_bar(fill = "coral", alpha = 0.7) +
    labs(
      title = paste("Distribution of", target_col),
      x = target_col,
      y = "Count"
    ) +
    theme_minimal()
  
  ggplotly(target_plot)
}
```

## Feature Distributions

```{r feature-distributions}
# Numerical features
numerical_features %
  select_if(is.numeric) %>%
  names()

if (length(numerical_features) > 0) {
  # Create distribution plots for numerical features
  numerical_plots %
    select(all_of(numerical_features)) %>%
    tidyr::pivot_longer(everything(), names_to = "variable", values_to = "value") %>%
    ggplot(aes(x = value)) +
    geom_histogram(bins = 30, fill = "skyblue", alpha = 0.7) +
    facet_wrap(~ variable, scales = "free") +
    labs(title = "Distribution of Numerical Features") +
    theme_minimal()
  
  print(numerical_plots)
}

# Categorical features
categorical_features %
  select_if(function(x) is.character(x) || is.factor(x)) %>%
  names()

if (length(categorical_features) > 0) {
  # Show top categories for each categorical feature
  for (cat_var in categorical_features[1:min(6, length(categorical_features))]) {
    cat_plot %
      count(!!sym(cat_var)) %>%
      top_n(10, n) %>%
      ggplot(aes(x = reorder(!!sym(cat_var), n), y = n)) +
      geom_col(fill = "lightgreen", alpha = 0.7) +
      coord_flip() +
      labs(
        title = paste("Top categories for", cat_var),
        x = cat_var,
        y = "Count"
      ) +
      theme_minimal()
    
    print(ggplotly(cat_plot))
  }
}
```

## Correlation Analysis

```{r correlation-analysis}
# Calculate correlations for numerical features
if (length(numerical_features) > 1) {
  cor_matrix %
    select(all_of(numerical_features)) %>%
    cor(use = "complete.obs")
  
  # Create correlation heatmap
  corrplot::corrplot(
    cor_matrix,
    method = "color",
    type = "upper",
    order = "hclust",
    tl.cex = 0.8,
    tl.col = "black",
    tl.srt = 45
  )
  
  # Interactive correlation heatmap
  plotly_cor %
    layout(title = "Feature Correlation Matrix")
  
  plotly_cor
}
```

## Outlier Detection

```{r outlier-detection}
# Box plots for numerical features
if (length(numerical_features) > 0) {
  boxplot_data %
    select(all_of(numerical_features)) %>%
    tidyr::pivot_longer(everything(), names_to = "variable", values_to = "value")
  
  outlier_plot %
    ggplot(aes(x = variable, y = value)) +
    geom_boxplot(fill = "orange", alpha = 0.7) +
    facet_wrap(~ variable, scales = "free") +
    labs(title = "Box Plots for Outlier Detection") +
    theme_minimal() +
    theme(axis.text.x = element_blank())
  
  print(outlier_plot)
}
```

## Data Quality Issues

```{r data-quality}
# Identify potential data quality issues
quality_issues  0) {
  quality_issues$duplicates %
  summarise_all(~ n_distinct(., na.rm = TRUE)) %>%
  tidyr::pivot_longer(everything(), names_to = "variable", values_to = "unique_values") %>%
  filter(unique_values <= 1) %>%
  pull(variable)

if (length(constant_cols) > 0) {
  quality_issues$constant_columns %
  select_if(function(x) is.character(x) || is.factor(x)) %>%
  summarise_all(~ n_distinct(.)) %>%
  tidyr::pivot_longer(everything(), names_to = "variable", values_to = "unique_values") %>%
  filter(unique_values > 0.9 * nrow(raw_data)) %>%
  pull(variable)

if (length(high_cardinality) > 0) {
  quality_issues$high_cardinality  0) {
  cat("Data Quality Issues Found:\n")
  for (issue in names(quality_issues)) {
    cat("-", quality_issues[[issue]], "\n")
  }
} else {
  cat("No major data quality issues detected.\n")
}
```

## Key Findings

```{r key-findings}
# Summarize key findings from exploration
findings  0), "variables with missing values"),
  "Numerical features" = length(numerical_features),
  "Categorical features" = length(categorical_features)
)

findings_df <- data.frame(
  Finding = names(findings),
  Description = unlist(findings)
)

knitr::kable(findings_df, caption = "Key Exploration Findings")
```

## Next Steps

Based on the exploration, the following steps are recommended:

1. **Data Cleaning**: Address missing values and outliers
2. **Feature Engineering**: Create new features based on domain knowledge
3. **Data Preprocessing**: Scale numerical features and encode categorical variables
4. **Model Selection**: Choose appropriate algorithms based on the data characteristics

---

*Analysis completed on `r Sys.Date()`*